Complexity-stability relationships disordered dynamical systems

### 2024-06-12 papers/response-referee.md

## Referee A

We thank the referee for the thoughtful review of our work and for raising important points, addressed below, which we think helped us improve the paper.

- *The claim is probably true, but unfortunately it does not describe a generic situation, because it is very hard to imagine that the exponents alpha, beta, gamma used in the model are identical for all populations. Perhaps the model would be more interesting if it included a spectrum of different exponents, described, for example, by a probabilistic law.*

We agree that indeed a spectrum of different exponents is probably more realistic, and thank the reviewer for pointing that out.

@@

We can consider the case in which the exponents are not identical for every variable $x_i$ but instead we have them from Gaussian distributions: $\alpha_i\sim\mathcal N(\alpha,\sigma_e)$, $\beta_i\sim\mathcal N(\beta,\sigma_e)$ and $\gamma_i\sim\mathcal N(\gamma,\sigma_e)$, respectively with mean $\alpha$, $\beta$ and $\gamma$, and, for simplicity, all with the same standard deviation $\sigma_e$.

Simulations show that all our results are robust as long as $\sigma_e$ is small enough that $\displaystyle{\min_i\beta_i>\max_i\alpha_i}$.

As an example, we show in the plot below the probability of stability vs. $\sigma_e$ for a system with $S=100$, $\mu=\mu_s=\sigma=0.01$, $\alpha=\gamma=1$ and $\beta=3/2$.
The probability of stability is obtained as the fraction of stable systems out of $100$ realizations for each value of $\sigma_e$.
![](../plots/stoch-exp.png)

We decided, however, to not include this analysis in the letter because we would like to keep the focus on the existence of the two complexity-stability regimes rather than exploring more detailed ecological scenarios. We stick to the case in which $f$, $g$ and $h$ are independent of the variable index, following references [18] and [19].

@@

- *The way of writing the matrix elements in Equations (7) is somewhat confusing. For clarity, the Kronecker delta can be used on the right side of these equations.*

We thank the referee for this comment, we modified Equations (7) accordingly.

@@

- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>. The self-consistency condition requires that the probability distribution (12) has exactly the same expected values. This should be discussed in detail.*

We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarification.

In some cases, for example, $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations. However in potential cases of interest, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge. We now modified the discussion removing the Gaussian approximation and considering the introduction of a cut-off. The idea behind this cut-off can be described by the example below.

Consider, as an example that allows for a complete analytical treatment, the case of a power law distribution 
$$
P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
$$
defined from 1 to $\infty$ and with 
$$
\mathcal{Z}=\int_1^{\infty}dxx^{-\alpha}.
$$

Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
$$
\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
$$
where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
$$
\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
$$
with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\alpha=3/2$ we have 
$$
\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
$$
and therefore $\Lambda=N^2$. We have for the mean
$$
\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
$$
The result is plotted below, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

@@

     *The rationale behind the proposed generalization of the GLV model is claimed to be "natural" based on physical and biological grounds, specifically using power laws. Nevertheless, the literature extensively demonstrates species abundance growth being well-described by logistic equations or oscillatory predator-prey equations, that indeed are linear (except for the carrying capacity term). The justifications provided are weak, with the provided examples failing to convincingly represent the complex dynamics of natural systems, such as forest biomass distribution. If we think to a forest there are various heights of trees, and in general the biomass in not uniformly distributed, and thus also the growth is not necessary constrained by the surface. The sublinear scaling of the “production function”, f, if this is really the right conclusion based on the references of the manuscript, must be the result of a cooperative behavior of the interacting system and as such it should not appear as hand-added like the authors do. Furthermore, in reference 27 the sublinear scaling with k=3/4 refers to the scaling of the single individual growth rather than to a community.*

We understand the criticism of the referee, pointing out the need for better justification of models of this kind in ecology. Our goal in this letter, however, is not to propose a model for a specific ecological scenario. Rather, we employ this model for complex systems (following references [18] and [19]) to show in a minimal setting the existence of the two opposite complexity-stability relationships in disordered dynamical systems, possibly with applications beyond ecology.

Nevertheless, the examples we provide are indicative that models of this class can be relevant in ecology (see reference [17], where a discussion of possible mechanisms is present; even though admittedly more work is required in that direction, which is beyond the scope of our paper).

Finally, we are aware of the fact that reference [27] refers to the scaling of single individuals, indeed we mention it to introduce the scaling at the higher level of organization:

@@

"Biologically, the growth of organisms (populations of cells) has long been known to scale like $f (x) \sim x^k$ with $k\simeq 3/4$ [27], which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities [17, 28]".

- *Feasibility Condition Neglected:*

    *The authors do not seem to consider the condition for the feasibility of the fixed points (x\* >= 0), which is critical. This oversight, especially evident in Equations 11 and 12, i.e., Eq. 12 or its Gaussian approximation are normalized from 0 or from -infty to infty? I think this is a very relevant issue calling into question the appropriateness of the results.*

We thank the referee for raising this point.
We have now specified that the distribution in Eq.(11) is defined from zero to infinity and that in the specific case we analyze the system is always feasible. This happens because, given that $\alpha<\beta$, at low enough density each degree of freedom effectively decouples from the others and only feels self-interactions and therefore never crosses zero.
We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distributions when a realistic lower threshold beyond which a species can be considered extinct is present.

@@

- *Lack of Proper Scaling:*

    *There is no proper scaling of the interactions, so the thermodynamic limit is not well defined. And this is not discussed or compared with other models in the literature applying similar disputable choice (see E Mallmin, A Traulsen, S De Monte - arXiv preprint arXiv:2306.11031, 2023). Relatedly, the comparison with GLV with quenched disorder (e.g. “The properties of large random dynamical systems is often portrayed in the ( sigma sqrt(N) , mu N ) plane [25] is not appropriate as such model scales the interactions strengths with the system size.*

This is a delicate point and we thank the referee for raising it. This gives us the opportunity to elaborate more on the choice of not scaling the interactions.

As also discussed in reference [25] (end of "I. MODEL DEFINITION"), the choice of scaling or not the interaction depends on the ecological setting of interest. 
We believe that it is appropriate to fix the strength and heterogeneity of the interaction ($mean(A)=\mu$ and $var(A)=\sigma^2$) if our goal is to study the effects of changes in the number of degrees of freedom.
In this setting, while the full limit in which $N\to\infty$ leads to the homogeneous interaction case, the cavity approximation does a good job at describing the empirical distributions obtained from simulations as long as $\sigma\sqrt{N}$ is not negligible with respect to $\mu N$.

We now discuss this and point out that it amounts to consider a strong-interactions regime and compare with the reference mentioned by the referee (E Mallmin, A Traulsen, S De Monte *PNAS* 2024, now reference [35] in the manuscript) for an ecological example in which this choice is made to study chaotic turnover of species with broad abundance distributions.

@@

We are grateful to the referee for this comment, it helps to clarify the scope of our work. We have now specified everywhere in the manuscript, including Abstract, Introduction and Discussion, that we compare out results to the *competitive* Lotka-Volterra model and in general that our results are relevant for predominantly competitive systems.

- *Notation:*

     *Using S instead of N for species richness would align better with ecological conventions.*

We decided to keep $N$ to indicate the number of degrees of freedom because our goal is not to align with ecological literature, rather we would like to speak to the community of scientists working in complex systems. We made this choice in the first place to align with Robert May notation in reference [1].

@@

- *Discussion Quality:*

    *Given the issues raised, the discussion section of the manuscript needs substantial improvement to convincingly argue the paper's contributions to the field. In light of these concerns, before the authors submit their manuscript elsewhere, I recommend to undertake significant revisions to address the weaknesses in their modelling framework, the rationale behind their approach, and the ecological implications of their findings*

We are thankful for the critical review overall. We extended the discussion and improved the paper in different aspects following the suggestions of the referee.

### 2024-06-12 papers/response-referee.md

Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behavior of the sample mean 
$$
\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
$$
where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
$$
\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
$$
with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\alpha=3/2$ we have 
$$
\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
$$
and therefore $\Lambda=N^2$. We have for the mean
$$
\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
$$
The result is plotted below, alongside the sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

![](../plots/cut-off-justification.png)

## Referee B

We thank the referee for the thorough and critical review. We address each of the comments below point by point. We believe the results of our paper are now more transparent thanks to the work needed to address the points raised by the referee.

@@

"Biologically, the growth of organisms (populations of cells) has long been known to scale like $f (x) \sim x^k$ with $k\simeq 3/4$ [27], which can be understood in terms of hydrodynamic constraints on vascular and pulmonary networks. For reasons that are not currently understood, a similar pattern of growth appears to recur at the level of ecological communities [17, 28]".

- *Feasibility Condition Neglected:*

    *The authors do not seem to consider the condition for the feasibility of the fixed points (x\* >= 0), which is critical. This oversight, especially evident in Equations 11 and 12, i.e., Eq. 12 or its Gaussian approximation are normalized from 0 or from -infty to infty? I think this is a very relevant issue calling into question the appropriateness of the results.*

We thank the referee for raising this point.
We have now specified that the distribution in Eq.(11) is defined from zero to infinity and that in the specific case we analyze the system is always feasible. This happens because, given that $\alpha<\beta$, at low enough density each degree of freedom effectively decouples from the others and only feels self-interactions and therefore never crosses zero.
We also now point to reference [17] where a discussion on the feasibility of ecological systems described by this kind of distribution when a realistic lower threshold beyond which a species can be considered extinct is present.

@@

    *The justification for employing a Gaussian approximation is unclear, given that Equation 12 should be solved self-consistently. Ecologically, Gaussian species abundance distributions (SADs) are not meaningful due to the prevalence of heavy-tailed SADs in natural ecosystems.*

We thank the referee for this important comment, we now removed the Gaussian approximation in the discussion and clarified how to treat the case in which self-consistently computing the moment in Equation (12) can be problematic, as explained above in the answers to referee A.

- *Homogenous Case and Stability:*
     *The manuscript concludes that stability requires stronger self-interactions than cross-interactions when alpha equals beta. While this is true that self-interactions promote stability also for the Lotka-Volterra model, any negative mean species interaction (indicating mutualism) in the homogeneous case leads to instability, irrespective of system size. Thus, comparing this to the GLV model seems a stretch.*

We are grateful to the referee for this comment, it helps to clarify the scope of our work. We have now specified everywhere in the manuscript, including Abstract, Introduction and Discussion, that we compare our results to the *competitive* Lotka-Volterra model and in general that our results are relevant for predominantly competitive systems.

### 2024-06-12 papers/submission/beyond-may.tex

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in the light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are in fact two kinds of complexity-stability relationships in disordered dynamical systems dominated by competition:
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
    Our result shows that May's principle that "complexity begets instability" is not a general property of complex systems.
\end{abstract}

@@

%Random interactions: stability condition
\paragraph*{}
\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with scaled mean $\mu=\tilde{\mu}/N$ and standard deviation $\sigma=\tilde{\sigma}/\sqrt{N}$~\cite{bunin2017ecological}.
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{align}
    M_{ij} = - g(x_i^*)&h'(x_i^*)\psi(x_i^*)\delta_{ij}, \nonumber \\ 
    L_{ij} = g(x_i^*)\delta_{ij}, & \quad R_{ij} = h'(x_i^*)\delta_{ij}, 
\end{align}
where $\delta_{ij}$ is the Kronecker delta and $\psi(x)$ is the function defined in~\eqref{eq: psi}.
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{cavity.pdf}
    \caption{The equilibrium distribution for the case of power-law self- and cross-interactions is accurately reproduced by Eq.~\eqref{eq: dist general}.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=100$, $\mu=\mu_s=\sigma=10^{-2}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this to hold, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \tilde{\mu}/N$ (independently of $x$) and for large $N$ we recover the classical condition $\tilde{\sigma} = \sigma\sqrt{N} < \mu_s$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depend on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative random variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently. XXX feasibility XXX

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case, the distribution can be normalized but the first and second moments formally diverge. It is not uncommon, for example in ecology, to find empirical species abundance distributions characterized by heavy-tails, potentially with diverging moments.
This is problem can be overcome considering that we are dealing with large, but \emph{finite}, populations. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$.

@@

The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
In particular, we are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions is fixed. This choice, equivalently expressed as having $\tilde{\mu}$ and $\tilde{\sigma}^2$ increasing with $N$, reflects a strong-interaction regime and has been also employed, for example, to study ecological scenarios whith large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}.
If we keep the mean $\mu$ and standard deviation $\sigma$ of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane. In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.

@@

Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\tilde{\mu}=\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c\sqrt{N}$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

@@

A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on the sum of the other densities (weighted by interaction strength) is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where $F$ is an arbitrary function.
General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
All the models in this class are destabilized by diversity.
The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.

>>>>

\begin{abstract}
    Robert May famously used random matrix theory to predict that large, complex systems cannot admit stable fixed points. 
    However, this general conclusion is not always supported by empirical observation: from cells to biomes, biological systems are large, complex and, often, stable.
    In this paper, we revisit May's argument in light of recent developments in both ecology and random matrix theory. 
    Using a non-linear generalization of the competitive Lotka-Volterra model, we show that there are, in fact, two kinds of complexity-stability relationships in disordered dynamical systems:
    if self-interactions grow faster with density than cross-interactions, complexity is destabilizing; but if cross-interactions grow faster than self-interactions, complexity is stabilizing.
\end{abstract}

@@

%Random interactions: stability condition
\paragraph*{}
\emph{Random interactions: stability condition.---}We now consider the case of random interactions.
Specifically, we assume that interaction coefficients $A_{ij}$ are drawn independently from a distribution with mean $\mu$ and standard deviation $\sigma$.
We assume the diagonal elements $A_{ii}$ have a mean value $\mu_s$ (possibly different from $\mu$) and standard deviation $\sigma$.

@@

To use this result, we decompose the interaction matrix as $A = \mu\mathbf{1} + (\mu_s-\mu)I + S$,
with $\mathbf{1}$ the matrix with all entries equal to $1$ and $S$ a random matrix as defined above.
Up to the rank-one perturbation $\mu\mathbf{1}$ which does not affect stability properties \cite{Stone2018}, we can write the Jacobian~\eqref{eq: jac off-diag} as $J = M + LSR$ with diagonal matrices
\begin{align}
    M_{ij} = - g(x_i^*)&h'(x_i^*)\psi(x_i^*)\delta_{ij}, \nonumber \\ 
    L_{ij} = g(x_i^*)\delta_{ij}, & \quad R_{ij} = h'(x_i^*)\delta_{ij}, 
\end{align}
where $\delta_{ij}$ is the Kronecker delta and $\psi(x)$ is the function defined in~\eqref{eq: psi}.
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{cavity.pdf}
    \caption{The equilibrium distribution for the case of power-law self- and cross-interactions is accurately reproduced by Eq.~\eqref{eq: dist general}.
    For simulations (in green) we used $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=100$, $\mu=\mu_s=\sigma=10^{-2}$.}
    \label{fig: cavity sol.}
\end{figure}
Stability of the equilibrium requires that $\mathcal{D}$ be entirely contained in the left half-plane (all eigenvalues have negative real part). For this to hold, it is no longer sufficient that $\psi(x_i^*) > 0$ for each $i$: we must also have that $0\notin \mathcal{D}$, hence from~\eqref{eq: domain}:
\begin{equation}
    \sum_{i=1}^N \Big(\frac{\sigma}{\psi(x_i^*)}\Big)^{2}
    < 1.
    \label{eq: random-stability}
\end{equation}
In the GLV model, $\psi(x) = \mu_s - \mu$ (independently of $x$) and we recover the classical condition $\sigma\sqrt{N} < \mu_s - \mu$ \footnote{In fact, the same condition holds when $f$ and $g$ are power laws with the same exponent.}. In general, however, the dependence on equilibrium values $x_i^*$ (which in turn depends on $N$) does not cancel out, and one must gain information about the distribution of equilibrium values $P(x_i^*)$ to assess the stability condition~\eqref{eq: random-stability}.

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution that describes the ensemble of the stationary solutions of the system by means of a single representative random variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \mu N \langle x_*^{\gamma}\rangle + \sigma\sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
We are interested in the case in which the number of degrees of freedom $N$ is the only parameter changing in the system and the strength and heterogeneity of the interactions are fixed~\cite{kessler2015generalized,fried2016communities}. This choice reflects a strong-interaction regime and has been employed, for example, to study ecological scenarios with large differences in species abundance and chaotic turnover~\cite{mallmin2024chaotic}. As we show below, the cavity approximation does a good job at describing the effects of heterogeneity in the interactions as long as $\sigma \sqrt{N}$ is not negligible relative to $\mu N$, at which point we recover the homogeneous case.

The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \mu N \langle x_*^{\gamma}\rangle + \sigma \sqrt{N} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\sigma^2N \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \mu N \langle x_*^{\gamma}\rangle)^2}{2\sigma^2N\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently.

@@

Let us focus on the choice $\alpha=1$, $\beta=3/2$ and $\gamma=1$, with $A_s=0$.  
In this case the system is always feasible (see Ref.~\cite{Hatton2024} for a discussion on the inclusion of extinction thresholds at low density in an ecological setting) and the distribution can be normalized. However, the first and second moments formally diverge.
This is problem can be overcome considering that we are dealing with a large, but \emph{finite}, number of degrees of freedom. 
We are therefore interested in predicting the sample mean and variance for a given $N$. Consider the sample mean $\bar{x}_{N}\equiv\sum_{i}^N x_i /N$. We can approximate it with $\bar{x}_N \approx \langle x_* \rangle_{\Lambda} \equiv \int_0^{\Lambda}dx_*x_*P(x_*)$ where the cut-off $\Lambda$ depends on $N$ in such a way that $\int_{\Lambda}^{\infty}dx_*P(x_*)=1/N$. In orther words such that there is less 1 variable with density above $\Lambda$.

@@

The properties of large random dynamical systems are often portrayed in the $(\sigma \sqrt{N},\mu N)$ plane \cite{bunin2017ecological}.
If we keep the mean $\mu$ and standard deviation $\sigma$ of interactions fixed, an increase in the number of degrees of freedom $N$ moves the system along a square-root trajectory in that plane. In the GLV model, the boundary between the stable and unstable parameter regions is the horizontal line $\sigma_c\sqrt{N} = \mu_s - \mu N / N \approx \mu_s$ \cite{bunin2017ecological}, hence stability is never possible at large $N$.

@@

Finally, Fig.~\ref{fig: alpha-beta} shows results of simulations for $\sigma_c\sqrt{N}$ in the $(\alpha,\beta)$ plane, at fixed $\mu N$ and $\gamma = 1$. (Numerically, we define $\sigma_c$ as the value of $\sigma$ above which full stable coexistence has a probability lower than $50\%$.) In these simulations, we use $\mu_s = \mu$, which would never be stable in the GLV model. Fig.~\ref{fig: alpha-beta} illustrates the transition between a ``May" phase for $\alpha \geq \beta$ and an ``anti-May" phase for $\alpha < \beta$, showing that our analytical results are robust to heterogeneous interactions.

@@

A modification of our model~\eqref{dynamics} of general ecological relevance describing the case in which per capita growth rates depend only on a linear combination of the densities is $\dot{x_i}/x_i=1-F(\sum_jA_{ij}x_j)$, where the function $F$ characterizes the effects of competition.
General models of this kind are used, for example, to study the growth of a clonal population in the presence of a limiting factor~\cite{mazzolini2023universality}.
All the models in this class are destabilized by diversity.
The Jacobian matrix at equilibrium is $J_{ij}^*=-A_{ij}x_i^*F'(\sum_jA_{ij}x_j^*) \quad \forall i,j$ leading to the same simplification of the stability condition~\eqref{eq: random-stability} possible for GLV.
